Non-kissing and non-crossing complexes \\ for locally gentle algebras

### 2018-07-05 draft/accordionComplexSurfaces.tex
proof purity

% others
\newcommand{\fref}[1]{Figure~\ref{#1}} % reference figures
\newcommand{\ie}{\textit{i.e.}~} % id est
\newcommand{\eg}{\textit{e.g.}~} % exempli gratia
\newcommand{\Eg}{\textit{E.g.}~} % exempli gratia
\newcommand{\apriori}{\textit{a priori}} % a priori
\newcommand{\viceversa}{\textit{vice versa}} % vice versa
\newcommand{\versus}{\textit{vs.}~} % versus
\newcommand{\aka}{\textit{a.k.a.}~} % also known as
\newcommand{\perse}{\textit{per se}} % per se
\newcommand{\ordinal}{\textsuperscript{th}} % th for ordinals
\newcommand{\ordinalst}{\textsuperscript{st}} % st for ordinals
\definecolor{darkblue}{rgb}{0,0,0.7} % darkblue color
\definecolor{green}{RGB}{57,181,74} % green color
\definecolor{violet}{RGB}{147,39,143} % violet color
\newcommand{\red}{\color{red}} % red command
\newcommand{\green}{\color{green}} % green command
\newcommand{\darkblue}{\color{darkblue}} % darkblue command
\newcommand{\defn}[1]{\textsl{\darkblue #1}} % emphasis of a definition
\newcommand{\para}[1]{\medskip\noindent\textbf{#1.}} % paragraph
\renewcommand{\topfraction}{1} % possibility to have one page of pictures
\renewcommand{\bottomfraction}{1} % possibility to have one page of pictures
%\renewcommand\labelitemi{$\diamond$} % redefine itemize default symbol
\newcommand{\ex}{_{\textrm{exm}}} % examples
\newcommand{\pa}{_{\textrm{pa}}} % path

@@

\section*{Acknowledgements}
We thank Sibylle Schroll for discussions on geometric models for gentle algebras.

@@

@@

\begin{remark}
A quick disclaimer about accordion complexes and slalom complexes of arbitrary cellular dissections.
As stated in Proposition~\ref{prop:conditionsDualDissections}, a dissection~$\dissection$ with a face containing more than one boundary edge does not admit a dual cellular dissection~$\dissection\dual$.
This prevents to use the definitions given in this section.
Although its accordion complex (resp.~slalom complex) could still be defined as in~\cite{MannevillePilaud-accordion} (resp.~\cite{GarverMcConville}) for the disk, the resulting complexes would be join of smaller accordion complexes (resp.~slalom complexes) defined in this paper.
See \cite[Prop.~2.4]{MannevillePilaud-accordion} for a detailed statement in the case of the disk.
\end{remark}

@@

@@

\begin{lemma}\label{lem:nkCycles}
 Let~$\omega$ be a non-kissing walk that contains an oriented cycle~$c$.
 Then~$\omega$ is of the form~$\sigma c^\infty$ or~$c^\infty\sigma$, for some substring~$\sigma$.
\end{lemma}

\begin{proof}
 Let~$c=\alpha_1\cdots\alpha_r$ be an oriented cycle in some walk~$\omega$.
 Assume that~$\omega$ is not of form~$\sigma c^\infty$ or~$c^\infty\sigma$.
 Then, up to cyclic reordering of the arrows in~$c$, it contains a substring of the form~$\beta^\pm c^l \alpha_1\cdots\alpha_j\gamma^\pm$, for some arrows~$\beta,\gamma$ not in~$c$, some~$l\geq 1$, and some~$j<r$.
 Since~$\bar Q$ is locally gentle, this substring is~$\beta^{-1} c^l \alpha_1\cdots\alpha_j\gamma^{-1}$.
 Hence, the walk~$\omega$ contains the two substrings~$\beta^{-1}\alpha_1\cdots\alpha_j\alpha_{j+1}$ and~$\alpha_r\alpha_1\cdots\alpha_j\gamma^{-1}$, and thus self-kisses.
\end{proof}

@@

\begin{lemma}
\label{lemm:countercurrentOrderIsTotal}
For any face~$F$ of~$\NKC$ and any arrow~$\alpha\in Q_1\blossom$, the countercurrent order~$\prec_\alpha$ defines a total order on~$F_{\alpha}$.
\end{lemma}
\begin{proof}
 The proof of \cite[Lemma 2.21]{PaluPilaudPlamondon} applies here.
\end{proof}

@@

 {\it Combinatorial proof.}
 Note that even though~$F$ is finite, the set of marked walks~$F_\alpha$ might be infinite if~$\alpha$ belongs to an oriented cycle of~$Q$.
 Assume that~$F_\alpha$ is non-empty, and infinite.
 Any walk that contains an infinite number of copies of~$\alpha$ can be written in the form~$\omega=\sigma c^\infty$ where~$\sigma$ is a (possibly infinite) substring and~$c$ is an oriented cycle containing~$\alpha$.
 Then~$\omega$ marked at the first occurrence of~$\alpha$ in~$c^\infty$ is greater than~$\omega$ marked at any other occurrence of~$\alpha$ in~$c^\infty$.
 We can thus safely remove all copies of~$\omega$ marked at an occurrence of~$\alpha$ in~$c^\infty$ but the first one.
 After applying this procedure a finite number of times,~$F_\alpha$ becomes finite, and the claim follows.

 {\it Geometric proof.} View~$F$ as a face of the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$.
 The arrow~$\alpha$ then corresponds to an angle in~$\dissection$ at a vertex~$v$.

@@

\begin{lemma}\label{lem:ExistenceOfDistinguishedArrows}
Let~$F$ be a finite face of~$\NKC$, let~$\omega$ be a walk in~$F$, and let~$\alpha$ be an arrow on~$\omega$.
Assume that~$\omega$ is not an infinite straight walk.
Then, there is an arrow~$\beta$ oriented in the same direction as~$\alpha$ in~$\omega$ such that~$\omega$ (marked at~$\beta$) is maximal in~$F$ for~$\prec_\beta$.
\end{lemma}

@@

\begin{proposition}
\label{prop:purity}
 The the non-kissing complex~$\NKC$ and the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ are pure of dimension~$3|Q_0| - |Q_1|$.
\end{proposition}

Note that Proposition \ref{prop:purity} was proved for~$\NKC$ in \cite[Corollary 2.27]{PaluPilaudPlamondon} in the case where~$kQ/I$ is finite-dimensional.
Note also that, by Theorem \ref{thm:complexesCoincide} above,~$\NKC$ and~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ coincide, so it suffices to prove the Proposition for one of them.

@@

 {\it Combinatorial proof.} By Lemma~\ref{lem:ExistenceOfDistinguishedArrows}, the number of walks in a finite face of~$\NKC$ is bounded by~$|Q_1\blossom|+s_\infty$, where~$s_\infty$ is the number of infinite straight walks (i.e. the number of oriented cycles in~$Q$).

 {\it Geometric proof.} Let~$F$ be a face of the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$.
 Its elements are pairwise compatible~$\dissection$-accordions on the surface~$\surface(\bar Q)$.
 Thus~$F$ can be viewed as a partial tagged triangulation of~$\surface(\bar Q)$, by letting the~$\dissection$-accordions circling a puncture in (say) clockwise order be seen as ``notched'' arcs to said puncture, and the ones circling it in counter-clockwise orientation be ``un-notched'' arcs.
 It is known (see, for instance, \cite{FominShapiroThurston}) that any such partial tagged triangulation can be embedded into a full triangulation, and that triangulations contain a finite number of arcs.
 Thus~$F$ is finite, and is contained in a finite maximal face.
\end{proof}

@@

The following definition from \cite[Definition 2.23]{PaluPilaudPlamondon} extends to the setting of locally gentle algebras.

@@

The following result extends \cite[Proposition 2.26]{PaluPilaudPlamondon} to the case of locally gentle algebras.

\begin{proposition}
\label{prop:numberOfDistinguishedArrows}
Let~$F$ be a facet of~$\NKC$.
\begin{itemize}
 \item Each bending walk of~$F$ has exactly two distinguished arrows.
 \item Each finite straight walk of~$F$ has exactly one distinguished arrows.
 \item Each infinite straight walk of~$F$ has exactly zero distinguished arrows.
\end{itemize}
\end{proposition}
\begin{proof}
 \vincent{Vincent has the proof of this proposition.}
 The first two dots in this proposition are proven in Lemma~\ref{lem:ExistenceOfDistinguishedArrows}.
\end{proof}

@@

\begin{proof}[Proof of Proposition \ref{prop:purity}.]
\vincent{TO DO}
\end{proof}

@@

A simplicial complex is \defn{thin} if each of its codimension~$1$ faces is contained in exaclty two facets.
We state the following results, which generalize \cite[Proposition 2.31 and Corollary 2.33]{PaluPilaudPlamondon}, whose proofs apply withouth change.

>>>>

% others
\newcommand{\fref}[1]{Figure~\ref{#1}} % reference figures
\newcommand{\ie}{\textit{i.e.}~} % id est
\newcommand{\eg}{\textit{e.g.}~} % exempli gratia
\newcommand{\Eg}{\textit{E.g.}~} % exempli gratia
\newcommand{\apriori}{\textit{a priori}} % a priori
\newcommand{\viceversa}{\textit{vice versa}} % vice versa
\newcommand{\versus}{\textit{vs.}~} % versus
\newcommand{\aka}{\textit{a.k.a.}~} % also known as
\newcommand{\perse}{\textit{per se}} % per se
\newcommand{\ordinal}{\textsuperscript{th}} % th for ordinals
\newcommand{\ordinalst}{\textsuperscript{st}} % st for ordinals
\definecolor{darkblue}{rgb}{0,0,0.7} % darkblue color
\definecolor{green}{RGB}{57,181,74} % green color
\definecolor{violet}{RGB}{147,39,143} % violet color
\newcommand{\red}{\color{red}} % red command
\newcommand{\green}{\color{green}} % green command
\newcommand{\darkblue}{\color{darkblue}} % darkblue command
\newcommand{\defn}[1]{\textsl{\darkblue #1}} % emphasis of a definition
\newcommand{\para}[1]{\medskip\noindent\textit{#1.}} % paragraph
\renewcommand{\topfraction}{1} % possibility to have one page of pictures
\renewcommand{\bottomfraction}{1} % possibility to have one page of pictures
%\renewcommand\labelitemi{$\diamond$} % redefine itemize default symbol
\newcommand{\ex}{_{\textrm{exm}}} % examples
\newcommand{\pa}{_{\textrm{pa}}} % path

@@

\section*{Acknowledgements}
We thank Sibylle Schroll for discussions on geometric models for gentle algebras.
We are also grateful to Salvatore Stella for discussions on accordion complexes of dissections of surfaces.

@@

\begin{remark}
\label{rem:sizeBlossomingQuiver}
Note that the blossoming quiver has~$2|Q_0|-|Q_1|$ incoming blossoms and~$2|Q_0|-|Q_1|$ outgoing blossoms.
Therefore, it has $|Q_0\blossom| = |Q_0| + 2(2|Q_0|-|Q_1|) = 5|Q_0|-2|Q_1|$ vertices and~$|Q_1\blossom| = |Q_1| + 2(2|Q_0|-|Q_1|) = 4|Q_0|-|Q_1|$ arrows.
\end{remark}

@@

\begin{remark}
A quick disclaimer about accordion complexes and slalom complexes of arbitrary cellular dissections.
As stated in Proposition~\ref{prop:conditionsDualDissections}, a cellular dissection~$\dissection$ with a face containing more than one boundary edge does not admit a dual cellular dissection~$\dissection\dual$.
This prevents to use the definitions given in this section.
Although its accordion complex (resp.~slalom complex) could still be defined as in~\cite{MannevillePilaud-accordion} (resp.~\cite{GarverMcConville}) for the disk, the resulting complexes would be join of smaller accordion complexes (resp.~slalom complexes) defined in this paper.
See \cite[Prop.~2.4]{MannevillePilaud-accordion} for a detailed statement in the case of the disk.
\end{remark}

@@

This section explores combinatorial properties of the non-crossing and non-kissing complexes, showing in particular that these complexes are pure and thin.
The proof follows ideas of~\cite{McConville, GarverMcConville} already reinterpreted for non-kissing complexes of gentle algebras in~\cite{PaluPilaudPlamondon}.
However, the setting of locally gentle algebras studied in this paper requires a little more care.

@@

We first gather two technical lemmas on properties of oriented cycles of the quiver~$\bar Q$.
Note that oriented cycles were forbidden in the setting of gentle algebras studied in~\cite{PaluPilaudPlamondon}.

\begin{lemma}
\label{lem:nkCycles}
Let~$\omega$ be a non-kissing walk that contains an oriented cycle~$c$.
Then~$\omega$ is of the form~$\sigma c^\infty$ or~$c^\infty\sigma$, for some substring~$\sigma$.
\end{lemma}

\begin{proof}
Let~$c=\alpha_1\cdots\alpha_r$ be an oriented cycle in some walk~$\omega$.
Assume that~$\omega$ is not of form~$\sigma c^\infty$ or~$c^\infty\sigma$.
Then, up to cyclic reordering of the arrows in~$c$, it contains a substring of the form~$\beta^\pm c^l \alpha_1\cdots\alpha_j\gamma^\pm$, for some arrows~$\beta,\gamma$ not in~$c$, some~$l\geq 1$, and some~$j<r$.
Since~$\bar Q$ is locally gentle, this substring is~$\beta^{-1} c^l \alpha_1\cdots\alpha_j\gamma^{-1}$.
Hence, the walk~$\omega$ contains the two substrings~$\beta^{-1}\alpha_1\cdots\alpha_j\alpha_{j+1}$ and~$\alpha_r\alpha_1\cdots\alpha_j\gamma^{-1}$, and thus self-kisses.
\vincent{Should we also give here a geometric argument?}
\end{proof}

\begin{lemma}
\label{lem:walksThroughcycles}
For any oriented cycle~$c$ in~$\bar Q$, any maximal non-kissing face contains a walk of the form~$\sigma c^\infty$ or~$c^\infty\sigma$, for some substring~$\sigma$ distinct from~$c^\infty$.
\end{lemma}

\begin{proof}
We provide two proofs of the lemma.
One is combinatorial and applies to~$\NKC$, and the other is geometric and applies to~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$.

\para{Combinatorial proof}
Consider a maximal non-kissing face~$F \in \NKC$.
We first argue that $F$ contains a walk distinct from~$c^\infty$ containing at least one arrow of~$c$.
\vincent{TODO}

\para{Geometric proof}

\vincent{Should we also give here a geometric argument? Pierre-Guy, could you add that?}

@@

\begin{lemma}
\label{lemm:countercurrentOrderIsTotal}
For any face~$F$ of~$\NKC$ and any arrow~$\alpha\in Q_1\blossom$, the countercurrent order~$\prec_\alpha$ defines a total order on~$F_{\alpha}$.
\end{lemma}
\begin{proof}
 The proof of \cite[Lem.~2.21]{PaluPilaudPlamondon} applies here.
\end{proof}

@@

 \para{Combinatorial proof}
 Note that even though~$F$ is finite, the set of marked walks~$F_\alpha$ might be infinite if~$\alpha$ belongs to an oriented cycle of~$Q$.
 Assume that~$F_\alpha$ is non-empty, and infinite.
 Any walk that contains an infinite number of copies of~$\alpha$ can be written in the form~$\omega=\sigma c^\infty$ where~$\sigma$ is a (possibly infinite) substring and~$c$ is an oriented cycle containing~$\alpha$.
 Then~$\omega$ marked at the first occurrence of~$\alpha$ in~$c^\infty$ is greater than~$\omega$ marked at any other occurrence of~$\alpha$ in~$c^\infty$.
 We can thus safely remove all copies of~$\omega$ marked at an occurrence of~$\alpha$ in~$c^\infty$ but the first one.
 After applying this procedure a finite number of times,~$F_\alpha$ becomes finite, and the claim follows.

 \para{Geometric proof}
 View~$F$ as a face of the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$.
 The arrow~$\alpha$ then corresponds to an angle in~$\dissection$ at a vertex~$v$.

@@

\begin{lemma}\label{lem:ExistenceOfDistinguishedArrows}
Let~$F$ be a finite non-kissing face of~$\NKC$, let~$\omega$ be a walk in~$F$, and let~$\alpha$ be an arrow on~$\omega$.
Assume that~$\omega$ is not an infinite straight walk.
Then, there is an arrow~$\beta$ oriented in the same direction as~$\alpha$ in~$\omega$ such that~$\omega$ (marked at~$\beta$) is maximal in~$F$ for~$\prec_\beta$.
\end{lemma}

@@

\begin{proposition}
\label{prop:purity}
The non-kissing complex~$\NKC$ and the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ are pure of dimension~$3|Q_0| - |Q_1|$.
\vincent{This is wrong. We need to count the infinite straight walks. I propose the next statement instead.}
\end{proposition}

\begin{proposition}
\label{prop:purity}
The reduced non-kissing complex~$\RNKC$ and the reduced non-crossing complex~$\RNCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ are pure of dimension~$|Q_0|$.
The non-kissing complex~$\NKC$ and the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ are pure of dimension~$3|Q_0| - |Q_1| + p$ where~$p$ is the number of oriented cycles in~$\bar Q$ or equivalently the number of punctures in~$\surface_{\bar Q}$.
\end{proposition}

Note that Proposition \ref{prop:purity} was proved for~$\NKC$ in \cite[Cor.~2.27]{PaluPilaudPlamondon} in the case where~$kQ/I$ is finite-dimensional.
Note also that, by Theorem \ref{thm:complexesCoincide} above,~$\NKC$ and~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$ coincide, so it suffices to prove the Proposition for one of them.

@@

 \para{Combinatorial proof}
 By Lemma~\ref{lem:ExistenceOfDistinguishedArrows}, the number of walks in a finite face of~$\NKC$ is bounded by~$|Q_1\blossom|+s_\infty$, where~$s_\infty$ is the number of infinite straight walks (i.e. the number of oriented cycles in~$Q$).

 \para{Geometric proof}
 Let~$F$ be a face of the non-crossing complex~$\NCC[\dissection_{\bar Q}, {\dissection_{\bar Q}}\dual]$.
 Its elements are pairwise compatible~$\dissection$-accordions on the surface~$\surface(\bar Q)$.
 Thus~$F$ can be viewed as a partial tagged triangulation of~$\surface(\bar Q)$, by letting the~$\dissection$-accordions circling a puncture in (say) clockwise order be seen as ``notched'' arcs to said puncture, and the ones circling it in counter-clockwise orientation be ``un-notched'' arcs.
 It is known (see, for instance, \cite{FominShapiroThurston}) that any such partial tagged triangulation can be embedded into a full triangulation, and that triangulations contain a finite number of arcs.
 Thus~$F$ is finite, and is contained in a finite maximal face.
\end{proof}

@@

The following definition from \cite[Def.~2.23]{PaluPilaudPlamondon} extends to the setting of locally gentle algebras.

@@

The following result extends \cite[Prop.~2.26]{PaluPilaudPlamondon} to the case of locally gentle algebras.

\begin{proposition}
\label{prop:numberOfDistinguishedArrows}
In a non-kissing facet~$F$ of~$\NKC$,
\begin{itemize}
 \item each bending walk of~$F$ has exactly two distinguished arrows,
 \item each finite straight walk of~$F$ has exactly one distinguished arrow,
 \item each infinite straight walk of~$F$ has no distinguished arrows.
\end{itemize}
\end{proposition}

\begin{proof}
The proof of \cite[Prop.~2.26]{PaluPilaudPlamondon} applies here to show that each walk in the non-kissing facet~$F$ has at most one distinguished arrow in each direction.
Lemma~\ref{lem:ExistenceOfDistinguishedArrows} therefore shows the first two points.
Finally, Lemma~\ref{lem:walksThroughcycles} shows the last point.
\end{proof}

@@

\begin{proof}[Proof of Proposition \ref{prop:purity}.]
The proof follows that of~\cite[Cor.~2.27]{PaluPilaudPlamondon}.
Let~$b$ denote the number of bending walks, $s$ the number of finite straight walks, and $p$ the number of infinite straight walks in a non-kissing facet~$F$ of~$\NKC$.
Since each finite straight walk uses two blossom arrows, we have~$s = 2|Q_0|-|Q_1|$ (see Remark~\ref{rem:sizeBlossomingQuiver}).
Moreover, the infinite straight walks are just the oriented cycles in~$\bar Q$.
Using Proposition~\ref{prop:numberOfDistinguishedArrows}, we obtain by double-counting that
\[
2 * b + 1 * s + 0 * c= |Q_1\blossom| = 4|Q_0|-|Q_1|
\]
from which we derive that~$b = |Q_0|$ and~$b+s+c = 3|Q_0|-|Q_1|+p$.
\end{proof}

@@

A simplicial complex is \defn{thin} if each of its codimension~$1$ faces is contained in exaclty two facets.
We state the following results, which generalize \cite[Prop.~2.31 \& Cor.~2.33]{PaluPilaudPlamondon}, whose proofs apply withouth change.
